
```{r setup_solutions, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise Solutions by Module


## Module 1

### General exercises

**Exercise:** Which of the following will give a different output from the other 3? Hint: Look at the help file for the `log()` function.

  1. `log(x=6, base=4)`
  
  2. `log(4, 6)`
  
  3. `log(base=4, x=6)`
  
  4. `log(6, 4)`

*Solution:*
`?log`, 2. `log(4,6)`

**Exercise:** Create a vector that goes from 5 to 25 by increments of 1.

*Solution:* `5:25` (or some variation)

**Exercise:** Create another vector that goes from 5 to 25 by increments of 1, but using a different method. Call this vector `V1`.

*Solution:* `V1 <- seq(5,25,by=1)` (or some variation)

**Exercise:** Extract the 5th element of `V1`.

*Solution:* `V1[5]`

**Exercise:** Remove the 7th element from the vector `V1` and call this new vector `V2`.

*Solution:* `V2 <- V1[-7]`

**Exercise:** How many elements in your vector, `V1`, are greater than 11?

*Solution:* `sum(V1>11)` = 14

**Exercise:** Create a matrix that contains the sequence of numbers from 1 to 16, going up in increments of 1. Let the matrix have 4 rows and have the matrix elements fill by row. Call this matrix `M1`.

*Solution:* `M1 <- matrix(data=1:16, nrow=4, byrow=T)`

**Exercise:** Extract the 3rd and 4th rows and the 1st and 2nd columns of your matrix, `M1`. Call this matrix `M2`.

*Solution:* `M2 <- M1[c(3,4), c(1,2)]`

**Exercise:** Find the row sums of the matrix `M2`.

*Solution:* `apply(M2, 1, sum)` (or `rowSums(M2)`)


### End of module exercises

**Exercise 1:** Use R to find the value of the square root of 25. ($\sqrt(25)$)
Hint: Look at the help file for the function `sqrt()`. 

*Solution 1:* `sqrt(25)`

**Exercise 2:** Use R to find the value of the exponential of $6\times 14 -3$. ($\exp(6\times 14-3)$)

*Solution 2:* `exp(6*4-3)`

**Exercise 3:** Use R to find the value of the absolute value of $7\times 3 - 4\times 9 - 30$. ($|7 \times 3 - 4 \times 9 - 30|$)

*Solution 3:* `abs(7*3-4*9-30)`

**Exercise 4:**  Use R to find the value of $\frac{73-42}{3}+2\times\left(\frac{36}{4}-17\right)$. Give your answer to 2 decimal places.

*Solution 4:* `(73-42)/3+(2*((36/4)-17))`

**Exercise 5:** How many elements does the following vector have?
  
  `seq(from = 0, to = 100, by = 3.14)`

*Solution 5:* `length(seq(0,100,by=3.14))`

**Exercise 6:**  Look at the help file for the function `rnorm()` and use this function to generate 100 random numbers with mean 10 and standard deviation 5.

*Solution 6:* `rnorm(n = 100, mean = 10, sd = 5)`




## Module 2

### General exercises

**Exercise:** What are the defaults for the header argument in the functions `read.table()` and `read.csv()`?

-   `header = T` in `read.table()` and `header = T` in `read.csv()`

-   `header = T` in `read.table()` and `header = F` in `read.csv()`

-   `header = F` in `read.table()` and `header = T` in `read.csv()`

-   `header = F` in `read.table()` and `header = F` in `read.csv()`

*Solution:* `header = F` in `read.table()` and `header = T` in `read.csv()`

### End of module exercises

**Exercise 1:** How many cares have a miles per gallon (`mpg`) of less than 15 in the `mtcars` data?

*Solution 1:*

`library(datasets)`

`cars <- mtcars`

`sum(cars$mpg<15)` = 5

**Exercise 2:** How many cars have exactly 4 cylinders (`cyl`) in the `mtcars` data?

*Solution 2:* `sum(cars$cyl==8)`=14

**Exercise 3:** What is the mean value of horsepower (`hp`) to 2 decimal places in the `mtcars` data?

*Solution 3:* `round(mean(cars$hp), 2)`=146.69

**Exercise 4:** What car has the lowest miles per gallon (`mpg`) value for cars with 8 cylinders (`cyl`) in the `mtcars` data?

*Solution 4:* `cars[which.min(cars$mpg),]`=Cadillac Fleetwood

**Exercise 5:** What is the median miles per gallon (`mpg`) value for cars with 8 cylinders (`cyl`) in the `mtcars` data?

*Solution 5:* `median(cars$mpg[cars$cyl==8])`= 15.2

**Exercise 6:** What car has the highest weight (`wt`) for each amount of cylinders (`cyl`) in the `mtcars` data?

*Solution 6:*

`unique(cars$cyl`

`cars[which.max(cars$wt[cars$cyl==4]),]`=Mazda RX4 Wag

`cars[which.max(cars$wt[cars$cyl==6]),]`=Hornet 4 Drive

`cars[which.max(cars$wt[cars$cyl==8]),]`=Duster 360

**Exercise 7:** Create a bar plot that shows the number of cars with each gear type (`gear`) in the `mtcars` data.

*Solution 7:*

`gear_types <- table(cars$gear)`

`barplot(gear_types, names.arg=c("Type 3", "Type 4", "Type 5"),`

`xlab = "Gear types", ylab = "Total count",`

`main = "Distribution of gear types")`

**Exercise 8:** Create a bar plot that shows the number of cars with each gear type (`gear`) with the distribution of cylinders (`cyl`) for each type in the `mtcars` data. Add a legend for the cylinders.

*Solution 8:*

`cyl_gear_types <- table(cars$cyl, cars$gear)`

`colours <- c("#004C92", "#FF0000", "#FFC600")`

`barplot(cyl_gear_types, names.arg=c("Type 3", "Type 4", "Type 5"),`

`xlab = "Gear types", ylab = "Total count",`

`main = "Distribution of gear types vs cylinders",`

`col = colours)`

`legend("topright", rownames(cyl_gear_types), cex = 1.3, fill = colours)`

**Exercise 9:** Create a scatter plot to show the relationship between weight (`wt`) and miles per gallon (`mpg`) in the `mtcars` data.

*Solution 9:*

`plot(x = cars$wt,y = cars$mpg,`

`xlab = "Weight",`

`ylab = "Miles per gallon",`

`xlim = c(1,6),`

`ylim = c(10,35),`

`main = "Weight vs miles per gallon")`

`ggplot(cars, aes(x = wt, y = mpg)) +`

`geom_point(col="#004C92")+`

`labs(x="Weight", y="Miles per gallon")+`

`ggtitle("Weight vs miles per gallon")`

**Exercise 10:** Create a scatter plot to show the relationship between weight (`wt`) and miles per gallon (`mpg`) by cylinder (`cyl`) in the `mtcars` data.

*Solution 10:*

`plot(cars$wt, cars$mpg, col = cars$cyl,`

`xlab="Weight", ylab="Miles per gallon",`

`main="Weight vs miles per gallon by cylinder")`

`lapply(cars$cyl, function(x) {`

`abline(lm(mpg ~ wt, cars, subset = (cyl == x)), col = x)})`

`legend("topright", rownames(cyl_gear_types),`

`col = c(4,6,8), pch = 1, bty = "n")`

`ggplot(cars, aes(x = wt, y = mpg, group=cyl, col=cyl)) +`

`geom_point() +`

`geom_smooth(method="lm", se=FALSE)+`

`labs(x="Weight", y="Miles per gallon")+`

`ggtitle("Weigth vs miles per gallon by cylinder")`


## Module 3
### General exercises

**Exercise:** How can you add multiple layers, such as polygons for states and points for health facilities, to a single map using `tmap`?

  1. By using `tm_shape()` and `tm_fill()`
  
  2. By using `tm_shape()` and `tm_dots()`
  
  3. By using multiple `tm_shape()` functions with different elements
  
  4. By using `tm_shape()` and `tm_lines()`

*Solution:*
3. By using multiple `tm_shape()` functions with different elements

### End of module exercises

**Exercise 1:** What is the primary purpose of the `.prj` file in a shapefile set, and which package is commonly used to handle this file type in R?

  1.  Stores map projection information, handled by `raster` package.
  
  2.  Stores attribute information, handled by `sp` package
  
  3.  Stores projection information, handled by `sf` package

  4.  Stores metadata information, handled by `terra` package

*Solution 1:* 3. Stores projection information, handled by `sf` package

**Exercise 2:** Create a map of Nigeria that combines the raster layer for building count and the vector layer for region.

*Solution 2:*
```{r, eval=F}
library(terra)     
library(sf)       
library(tmap)     

#load data
building_raster <- rast(nga_building_count.tif) #raster data
regions_vector <- st_read(nga_regions.shp) #vector data

#both are in same CRS
if (st_crs(regions_vector)$epsg != crs(building_raster, describe=TRUE)$code) {
 regions_vector <- st_transform(regions_vector, crs(building_raster))
}
 
#plot tmap
tmap_mode("plot")
tm_shape(building_raster) +
 tm_raster(style = "quantile", palette = "YlOrRd", title = "Building Count") +
tm_shape(regions_vector) +
 tm_borders(lwd = 1, col = "black") 
```

**Exercise 3:** Create a choropleth map that shows the variation in population density for Nigeria at region level. Export the results as a .png file.

*Solution 3:*
```{r, eval=F}
library(sf)         
library(tmap)       
library(dplyr)     
 
#load data
nigeria_regions <- st_read(nga_regions.shp)
 
#calculate population density
nigeria_regions <- nigeria_regions %>%
 mutate(pop_density = nga_pop / nga_area)
 
#create choropleth map
tmap_mode("plot")
choropleth_map <- tm_shape(nigeria_regions) +
tm_polygons("pop_density",
             style = "quantile",     
             palette = "Blues") +
 tm_layout(title = "Nigeria: Population Density by Region")
 
#save as PNG
tmap_save(choropleth_map, filename = "nigeria_population_density.png", 
          width = 2000, height = 1500, dpi = 300)
```

**Exercise 4:** How do you plot a raster dataset using `tmap`, ensuring it can handle large datasets efficiently?

1.  `tm_shape(raster_data)` + `tm_fill()`
2.  `tm_shape(raster_data)` + `tm_raster()`
3.  `tm_shape(raster_data)` + `tm_polygons()`
4.  `tm_shape(raster_data)` + `tm_dots()`

*Solution 4:*
2. `tm_shape(raster_data)` + `tm_raster()`


## Module 4

### General exercises

**Exercise:** Identify the dependent and independent variables in the following sentence:

'A researcher investigates the effects of school density on the grades of its pupils.'

*Solution:* Dependent variable: the grades of the pupils, independent variable: school density

**Exercise:**  Fit a Poisson GLM using the `ccancer` dataset exploring the relationship between the count of
cancer deaths and the covariates for the cancer site and region in Canada. Is the relationship between the
dependent and independent variables significant?

*Solution:* 
`ccancer_glm_exercise <- glm(Count ~ Site + Region, family = "poisson", data = ccancer)`

`summary(ccancer_glm_exercise)` 

Yes, the relationships are significant.

**Exercise:** What is the expected birth weight for a baby boy at a gestational age of 39.5 weeks?

*Solution:* $$Weight = -1447.24 -163.04\times 0+120.89\times 39.5 = 3327.92g$$

## Module 5

### General exercises

**Exercise:** Following on from the example above, what is the probability that an individual that is vaccincated but has a high fever is infected with disease A?

*Solution:* 0.19

**Exercise:** Following on from the example above, what is the conditional probability of drawing a purple button given that the button drawn is not blue?

*Solution:* 

$$
\begin{aligned}
P(A|B)&=\frac{P(A\cap B)}{P(B)}\\
&= \frac{\text{number of pruple buttons}/\text{total number of buttons}}{\text{blue}/\text{total number of buttons}}\\
&=\frac{4/10}{6/10}\\
&=\frac{2}{3}=0.67.
\end{aligned}
$$


**Exercise:** Following on from the example above, what would the probability be of selecting 2 men from the island without replacement?

*Solution:* 
$$
\begin{aligned}
P(M_1 \cap M_2)&=P(M_1)P(M_2|M_1)\\
&= \frac{\text{number of men}}{\text{total number of people}}\times\frac{\text{number of men left}}{\text{total number of people left}}\\
&=\frac{4}{10}\times\frac{3}{9}\\
&=\frac{2}{15}.
\end{aligned}
$$
## Module 6

### General exercises

**Exercise:** Following on from the example above, what is the probability that someone who is vaccinated is from region C?

*Solution:* 
$$P(A|Vaccinated) = \frac{P(C)P(Vaccinated|C)}{P(Vaccinated)} = \frac{0.3 \times 0.74}{0.728} = 0.305$$


**Exercise:** Create a histogram of posterior values for hospital D.

*Solution:* 
```{r hisogram, eval=F}
#histogram of posterior values for hospital D
ggplot(data = predicted_values_df , aes(x = D)) +
  geom_histogram(bins = 20, color = "#00b4d8", alpha = 1, fill = "#1d3557") +
  theme_bw() +
  labs(title = "Posterior Predicted values for D",
       x = "Posteriors",
       y = "Frequency")
```


**Exercise:** Create a density plot of the posterior values for hospital G.

*Solution:* 
```{r density, eval=F}
#density plot of posterior values for hospital G
ggplot(data = predicted_values_df , aes(x = G)) +
  geom_density(alpha = 1, fill = "#1d3557") +
  theme_bw() +
  labs(title = "Posterior predicted Values for G",
       x = "Posteriors",
       y = "Density")
```
